Adaptive Sequential Model Selection

### 2015-05-08 paper.tex
more work on 'Sequential Hypothesis Testing' section, reworking some notation.

We will consider two related problems. First, we will consider the problem of obtaining selective single-step $p$-values, where $p_k$ is a $p$-value for testing
\[
H_{0,k}:\; F\in M_{k-1}\quad \text{ vs. } \quad 
H_{1,k}:\; F\in M_k\setminus M_{k-1},
\]
while adjusting for the fact that the models are chosen adaptively. As we will see, {\em selected-model} tests can be dramatically more powerful than {\em saturated-model} tests at early steps in the model path when most of the signal variables have not yet entered the model.

@@

@@

To begin, we consider the problem of constructing a valid selective $p$-value for a particular step. At step $k$, we construct $p_k(Y)$ to test
\[
  H_{0,k}:\; F \in M_{k-1}(Y)
  \quad \text{ vs. } \quad
  H_{1,k}:\; F \in M_k(Y) \setminus M_{k-1}(Y).
\]
The main complication arises from the fact that the null and alternative hypotheses are random. To simplify matters, we first consider deriving $p$-values for a fixed candidate pair $m_{k-1} \sub m_{k-1}$.

The random variable  $p_{k,M}(Y)$ is a valid {\em selective $p$-value} for the fixed candidate pair $(m_{k-1},m_k)$ if it is stochastically larger than uniform under the (fixed) null $m_{k-1}$, given that the pair $(m_{k-1},m_k)$ is selected. That is,
\[
\P_F\left(p_{k,M}(Y) \leq \alpha \mid M_{k-1}(Y) = m_{k-1}\right) 
\leq \alpha, \quad \forall F\in m_{k-1}, \alpha \in [0,1].
\]
Given selective $p$-values for each fixed candidate pair, we can construct a combined $p$-value for the random null $(M_{k-1},M_k)$: 
\[
p_k(y) = p_{k, M_{k-1}(y), M_k(y)}(y),
\]
which is a valid $p$-value on the event $\{F \in M_{k-1}(Y)\}$:
\[
\P_F\left(p_k \leq \alpha \mid M_{k-1}, \;M_k, 
  \;F\in M_{k-1}\right) \leq \alpha, \quad \forall \alpha \in [0,1].
\]
Recall that, as usual, $F$ is not random, but $M_{k-1}$ is.

@@

\subsection{Selected- and Saturated-Model inference}

@@

Having discussed methods for constructing single-step $p$-values, we turn now to the problem of constructing a {\em stopping rule}; that is, an estimator $\hk$ of the completion index $k_0(Y)\in \{0,\ldots,d,\infty\}$. $M_k$ is ``rejected'' if and only if $\hk>k$. Thus $\hk$ is the number of models we rejected, while $k_0$ is the number that we should have rejected. The number of type I errors, then, is $(\hk-k_0)_+$, while the number of type II errors is $(k_0-\hk)_+$. Define $V=(\hk-k_0)_+$, the number of excess rejections.

Note that the type I error $V$ is defined in a ``model-centric,'' as opposed to a ``variable-centric,'' fashion: in regression, our model-wise $V$ counts the number of true null {\em models} incorrectly rejected, instead of counting the number of noise variables included in the final model. We can define the (model-wise) familywise error rate (FWER) and false discovery rate (FDR) respectively as $\P_F(V >0)$ and $\E_F[V/(\hk \vee 1)]$.

@@

While strong stop and forward stop tend to give more powerful stopping rules, they require independence among the $p$-values. Sections~\ref{sec:modelSPSP}--\ref{sec:pvalSP} discuss conditions on the model sequence $M_{[d]}$ and the $p$-value sequence $p_{[d]}$ under which selective $p$-values are mutually independent. We will see that, typically, selected-model $p$-values are sequentially independent while saturated-model $p$-values are not.

@@

\subsection{$p$-Values and Independence}

@@

  If the $p$-values are exact then $p_{k+1}, \ldots, p_d$ are
  uniform given $\sF_{k_0}$ on the event $\{k_0=k\}$.

  If they $p$-values are conservative, then for
  $\alpha_{k+1},\ldots,\alpha_d \in [0,1]$,
  \begin{equation}\label{eq:jointConserv}
  \P_F\left(p_{k+1}\leq \alpha_{k+1}, \ldots, p_d \leq \alpha_d\;
    \mid\; k_0 \leq k, \; \sF_k\right) \leqAS \prod_{i=k+1}^d
  \alpha_i.
  \end{equation}
\end{proposition}
Equation~\ref{eq:jointConserv} is slightly weaker than independence, but it guarantees that both Strong Stop and Forward Stop control their target error rates.

@@

  If the $p_k$ are exact then the above 
  inequalities become equalities, implying uniformity.
\end{proof}

\subsection{Sufficient Filtration}

@@

In most cases of interest, however, the two requirements are equivalent. First, we typically construct $p_k$ by conditioning on $T_{k-1}$. In fact, $T_{k-1}$ is complete sufficient, then any exact test must condition on $T_{k-1}$ \WFcomment{should we bother proving this?}. Second, most path algorithms of interest (including forward stepwise regression, the ever-active LASSO path, and their generalizations) satisfy a {\em subpath sufficiency principle} (henceforth, SSP): that once we know $M_k$ and its sufficient statistics, we can reconstruct the model path up to step $k$. Thus, most single-step $p$-values are already conservative conditional on $\sF_{k-1}$.

@@

@@

\begin{model}[Generalized Sparse Model]\label{mod:genSparse}
  Let $M$ be a model parameterized by $\theta\in \Theta \sub \R^p$:
  \[
  M = \{F_\theta:\; \theta \in \Theta\}.
  \]
  For subsets $E\sub [p]$ define the submodel with active coefficients $E$ as follows:
  \[
  \Theta(E) = \{\theta:\; \theta_j = 0, \;\;\forall j \notin E\}, 
  \quad M(E) = \{F_\theta:\; \theta\in \Theta_E\}.
  \]
  Let $L(\theta; Y)$ denote the likelihood for model $M$.
\end{model}
Note that, although we will make no assumptions about the form of $M$, the results of this section are only useful in cases where we can actually construct single-step $p$-values. At present, that caveat largely restricts us to using exponential family models.

@@

\begin{example}[Principle Components Analysis]
\WFcomment{Do this one.}  
\end{example}

All forward stepwise procedures satisfy the SPSP.

\begin{proposition}[Forward Stepwise Satisfies SPSP]\label{prop:forwardSPSP}
  The forward stepwise algorithm in Model~\ref{mod:genSparse} as defined in (\ref{eq:forwardDef_start}--\ref{eq:forwardDef_end}) satisfies the SPSP.
\end{proposition}
\begin{proof}
  Let $A$ denote the event $\{M_k = M(E)\}$ for some fixed active set $E$. Letting $T(Y)$ be a sufficient statistic for $m_k$, the likelihood $L$ restricted to $\Theta(E) \times A$ is proportional to a function depending only on $T$. If $L_T$ is the likelihood of $T(Y)$, then
\[
L(\theta; y) = L_T(\theta; T(y)) \;L_{Y\mid T}(y), \;\; \text{ for } (\theta, y) \in  \Theta(E) \times A.
\]
Note that $L_{Y \mid T}(y)$ does not depend on $\theta$ since $T$ is sufficient for $M(E)$.

For $y\in A$ and $i \leq k$, we have $E_i(y) \sub E$. This means that the maximizer in \eqref{eq:forwardDef_start} is attained for some $j\in E$. So, for $i \leq k$, on $A$,
\begin{align}
  j_i &= \argmax_{j\in E} \;\;\sup \left\{L(\theta; Y):\;
    \theta\in\Theta(E_{i-1} \cup \{j\})\right\} \\
  &= \argmax_{j\in E} \;\;\sup \left\{L_T(\theta; T(Y)):\;
    \theta\in\Theta(E_{i-1} \cup \{j\})\right\}. \label{eq:onlyT}
\end{align}
Replacing $L$ with $L_T$ in~\eqref{eq:onlyT} is justified because $\Theta(E_{i - 1} \cup \{j\}) \sub \Theta(E)$ when $E_{i-1}\cup E$ and $j\in E$.

Equation~\eqref{eq:onlyT} shows that $j_1,\ldots, j_{k-1}$ all depend on $Y$ only through $T(Y)$. As a result, it also follows that the sequence $M_{[k-1]}(Y)$ depends only of $T(Y)$.
\end{proof}

Another class of model selection procedures satisfying the SPSP is the sequence ``ever-active'' sets in regularized likelihood methods. For $r=0,1,\ldots$, let $P_r(\theta)$ denote a regularizing penalty, and define
\begin{align}\label{eq:regPathDef_start}
  \hat\theta^{r}(Y) &= 
  \argmin_{\theta\in\Theta} -\log L(\theta; Y) + P_r(\theta) \\
  \tE_r(Y) &= \left\{j:\; \hat\theta_j^s \neq 0 
    \text{ for any } s \leq r \right\}
\end{align}

@@

\begin{proposition}[Regularized Likelihood Paths Satisfy SPSP]\label{prop:regPathSPSP}
Ever-active model paths derived from regularized likelihood methods as in (\ref{eq:regPathDef_start}--\ref{eq:regPathDef_end}) satisfy the sub-path sufficiency principle.
\end{proposition}

\begin{proof}
  Again, let $A$ denote the event $\{M_k = M(E)\}$ for some fixed $E$, and define $T$ and $L_T$ as in the proof of Proposition~\ref{prop:forwardSPSP}. We can partition $A$ into subsets $A_r = A \cap \{R(k)=r\}$.

On the event $A_r$, every $\hat\theta^s$ for $s\leq r$ is supported on $E$. That is, for $s \leq r$, we can write 
\begin{equation}\label{eq:onlyT_Ar}
  \hat\theta^{s} = 
  \argmin_{\theta\in\Theta(E)} -\log L_T(\theta; T(Y)) + P_s(\theta) \quad \text{ on } A_r
\end{equation}
Equation~\label{eq:onlyT_Ar} guarantees that $M_{[k-1]}$ is a function of $T(Y)$ on $A_r$. That is, we can write
\[
M_{[k-1]}(Y) = M_{[k-1]}^r(T(Y)) \quad \text{ on } A_r
\]
It remains only to show that, on $A$, $R_k(Y)$ depends only on $T(Y)$; then we have
\[
M_{[k-1]}(Y) = M_{[k-1]}^{R(T)}(T) \quad \text{ on } A
\]
But $R_k$ is a function of $T$ on $A$ for the same reason: that $\hat\theta^s$ depends only on $T$ until other coefficients enter the model.
\WFcomment{Would the proof be cleaner with $A_r$ defined as $A\cap \{R_k \leq r\}$?}
\end{proof}

\subsection{Independent $p$-Values Via Selection Variables}\label{sec:selectionVariables}

As discussed in \citet{fithian2014optimal}, there are occasions when it is desirable for some reason to condition on more than just the selection event. On these occasions we can introduce a {\em selection variable} --- any random variable that is finer than the indicator of the selection event --- and condition on its value. Conditioning on a finer selection variable preserves validity conditional on the selection event, but tends to reduce the power of a test.

For example, in linear regression after selecting variables via the lasso, \citet{lee2013exact} propose conditioning not only on the set of active variables but also on the signs of the active coefficients. The selection event corresponding to a given active set $E\sub [p]$ typically contains up to $2^{|E|}$ convex polytopes, each of which corresponds to a particular sign pattern $\hat z_E$. Thus, conditioning on the selection variable $\hat z_E$ can simplify computation considerably.

We say that $p_{k,M}$ is a selective $p$-value with respect to selection variable $S(Y)$ if it is uniform under $F\in m_k$ given $S$.

To achieve joint error control properties among different tests, we may need to condition on a finer variable.
\end{comment}

@@

\begin{comment}
Sections~\ref{sec:pvalSP}--\ref{sec:modelSPSP} discuss sufficient conditions on the $p$-values and the selection algorithm under which single-step $p$-values are automatically independent. Section~\ref{sec:selectionVariables} discusses how we can create independent $p$-values by conditioning on finer selection variables at each step.

>>>>

We will consider two related problems. First, we will consider the problem of obtaining selective single-step $p$-values, where $p_k$ is a $p$-value for testing
\[
H_{0,k}:\; F\in M_{k-1}\quad \text{ vs. } \quad 
H_{1,k}:\; F\in M_\infty\setminus M_{k-1},
\]
while adjusting for the fact that the models are chosen adaptively. As we will see, {\em selected-model} tests can be dramatically more powerful than {\em saturated-model} tests at early steps in the model path when most of the signal variables have not yet entered the model.

@@

Note that the type I error $V=(\hk-k_0)_+$ is defined in a ``model-centric,'' as opposed to a ``variable-centric,'' fashion: in regression, our model-wise $V$ counts the number of true null {\em models} incorrectly rejected, instead of counting the number of noise variables included in the final model. We can define the (model-wise) familywise error rate (FWER) and false discovery rate (FDR) respectively as $\P_F(V >0)$ and $\E_F[V/(\hk \vee 1)]$.

@@

To begin, we consider the problem of constructing a valid selective $p$-value for a particular step. At step $k$, we construct $p_k(Y)$ to test
\[
  H_{0,k}:\; F \in M_{k-1}(Y)
  \quad \text{ vs. } \quad
  H_{1,k}:\; F \in M_\infty \setminus M_{k-1}(Y).
\]
The main complication arises from the fact that the null and alternative hypotheses are random. To simplify matters, we first consider deriving $p$-values for a fixed candidate hypothesis $M\sub M_{\infty}$.

The random variable  $p_{k,M}(Y)$ is a valid {\em selective $p$-value} for $M$ if it is stochastically larger than uniform under sampling distributions $F\in M$, given that $M$ is selected. That is,
\[
\P_F\left(p_{k,M}(Y) \leq \alpha \mid M_{k-1}(Y) = M\right) 
\leq \alpha, \quad \forall F\in M, \alpha \in [0,1].
\]
Given selective $p$-values for each fixed candidate pair, we can construct a combined $p$-value for the random null $M_{k-1}(Y)$: 
\[
p_k(y) = p_{k, M_{k-1}(y)}(y),
\]
which is a valid $p$-value on the event $\{F \in M_{k-1}(Y)\}$:
\[
\P_F\left(p_k \leq \alpha \mid M_{k-1}, 
  \;F\in M_{k-1}\right) \leq \alpha, \quad \forall \alpha \in [0,1].
\]
Recall that, as usual $M_{k-1}$ is random and not $F$.

@@

\subsection{Comparison of Selected- and Saturated-Model inference}

@@

Having discussed methods for constructing single-step $p$-values, we turn now to the problem of constructing a {\em stopping rule}; that is, an estimator $\hk$ of the completion index $k_0(Y)\in \{0,\ldots,d,\infty\}$. $M_k$ is ``rejected'' if and only if $\hk>k$. Thus $\hk$ is the number of models we rejected, while $k_0$ is the number that we should have rejected. The number of type I errors, then, is $(\hk-k_0)_+$, while the number of type II errors is $(k_0-\hk)_+$.

@@

While strong stop and forward stop tend to give more powerful stopping rules, they require independence among the $p$-values. Sections~\ref{sec:pValsIndep}--\ref{sec:modelSSP} discuss conditions on the model sequence $M_{[d]}$ and the $p$-value sequence $p_{[d]}$ under which selective $p$-values are mutually independent. We will see that, typically, selected-model $p$-values are sequentially independent while saturated-model $p$-values are not.

@@

\subsection{$p$-Values and Independence}\label{sec:pValsIndep}

@@

  If the $p$-values are exact then $p_{k+1}, \ldots, p_d$ are
  independent and uniform given $\sF_{k_0}$ on the event $\{k_0=k\}$.
  If they are conservative, then for
  $\alpha_{k+1},\ldots,\alpha_d \in [0,1]$,
  \begin{equation}\label{eq:jointConserv}
  \P_F\left(p_{k+1}\leq \alpha_{k+1}, \ldots, p_d \leq \alpha_d\;
    \mid\; k_0 \leq k, \; \sF_k\right) \leqAS \prod_{i=k+1}^d
  \alpha_i.
  \end{equation}
\end{proposition}
Equation~\ref{eq:jointConserv} does not quite imply independence, but it guarantees that both Strong Stop and Forward Stop control their target error rates.

@@

  If the $p_k$ are exact then the above 
  inequalities become equalities, implying uniformity and joint independence.
\end{proof}

\subsection{The Sufficient Filtration}\label{sec:suffFilt}

@@

In most cases of interest, however, the two requirements are equivalent. First, we typically construct $p_k$ by conditioning on $T_{k-1}$. In fact, if $T_{k-1}$ is complete sufficient, then any exact test must condition on $T_{k-1}$ (apply the definition of completeness to the function $h(Y) = \P(p_k(Y) \leq \alpha \mid T_{k-1}(Y))$). Second, most path algorithms of interest (including forward stepwise regression, the ever-active LASSO path, and their generalizations) satisfy a {\em subpath sufficiency principle} (henceforth, SSP): that once we know $M_k$ and its sufficient statistics, we can reconstruct the model path up to step $k$. Thus, most single-step $p$-values are already conservative conditional on $\sF_{k-1}$.

@@

\subsection{Model Paths and the Subpath Sufficiency Principle}\label{sec:modelSSP}

In Section~\ref{sec:suffFilt} we introduced the sufficient filtration as a way to get separated $p$-values. Because conditioning on $(M_{[k]}, T_k)$ is nominally different from conditioning only on $(M_k, T_k)$, we might suspect that we will sacrifice power by requiring $p_k$ to condition on the former.

@@

\begin{model}[Generalized Sparse Model]\label{mod:genSparse}
  Let $M_\infty$ be a model parameterized by $\theta\in \Theta \sub \R^p$:
  \[
  M_\infty = \{F_\theta:\; \theta \in \Theta\}.
  \]
  For subsets $E\sub [p]$ define the submodel with active coefficients $E$ as follows:
  \[
  \Theta(E) = \{\theta:\; \theta_j = 0, \;\;\forall j \notin E\}, 
  \quad M(E) = \{F_\theta:\; \theta\in \Theta_E\}.
  \]
  Let $\ell(\theta; Y)$ denote the log-likelihood for model
  $M_\infty$, and define the restricted log-likelihood
  \[
  \ell_E(\theta; Y) = \left\{\begin{matrix} \ell(\theta; \;Y) 
      & \theta \in \Theta(E)\\ -\infty  & 
      \mathrm{ otherwise.}\end{matrix}\right. 
  \]
\end{model}
Although we will make no formal assumptions about the form of $M_\infty$, the results of this section are only useful in cases where we can actually construct single-step $p$-values. At present, that caveat  largely restricts us to using exponential family models.

@@

\begin{example}[Principle Components Analysis]
\WFcomment{Do this one. Actually need infinitely many $\theta$s for this.}  
\end{example}

All forward stepwise procedures satisfy the SSP.

\begin{proposition}[Forward Stepwise Satisfies SSP]\label{prop:forwardSSP}
  The forward stepwise algorithm in Model~\ref{mod:genSparse} as defined in (\ref{eq:forwardDef_start}--\ref{eq:forwardDef_end}) satisfies the SSP.
\end{proposition}
\begin{proof}
  Let $A$ denote the event $\{M_k = M(E)\}$ for some fixed active set $E$. Writing $T(Y) = T(Y; \;M(E))$, the likelihood restricted to $\Theta(E)$ is proportional to a function depending only on $T$:
\[
\ell(\theta; Y) = \ell_E^T(\theta; T(Y)) + \ell_E^{Y \mid T}(Y, T(Y)),
\]
where the second term does not depend on $\theta$ because $T$ is sufficient.

On $A$, we have $E_i(Y) \sub E$ for all $i=1,\ldots, k$, meaning that the maximum in \eqref{eq:forwardDef_start} was attained by some $j\in E$ at every step. So, for $i \leq k$, we have
\begin{align}
  j_i &= \argmax_{j\in E} \;\;\sup \left\{\ell_E(\theta; Y):\;
    \theta\in\Theta(E_{i-1} \cup \{j\})\right\} \\
  &= \argmax_{j\in E} \;\;\sup \left\{\ell_E^T(\theta; T(Y)):\;
    \theta\in\Theta(E_{i-1} \cup \{j\})\right\}. \label{eq:onlyT}
\end{align}

Equation~\eqref{eq:onlyT} shows that $j_1,\ldots, j_{k-1}$ all depend on $Y$ only through $T(Y)$. As a result, it also follows that the enitre sequence $M_{[k]}(Y)$ depends only on $T(Y)$.
\end{proof}

Another class of model selection procedures satisfying the SSP is the sequence ``ever-active'' sets in any regularized likelihood method. For $r=0,1,\ldots,m$, let $P_r(\theta)$ denote a regularizing penalty, and define
\begin{align}\label{eq:regPathDef_start}
  \hat\theta^{r}(Y) &= 
  \argmin_{\theta\in\Theta} -\ell(\theta; Y) + P_r(\theta) \\
  \tE_r(Y) &= \left\{j:\; \hat\theta_j^s \neq 0 
    \text{ for any } s \leq r \right\}
\end{align}

@@

\begin{proposition}[Regularized Likelihood Paths Satisfy SSP]\label{prop:regPathSSP}
Ever-active model paths derived from regularized likelihood methods as in (\ref{eq:regPathDef_start}--\ref{eq:regPathDef_end}) satisfy the sub-path sufficiency principle.
\end{proposition}

\begin{proof}
  Again, let $A$ denote the event $\{M_k = M(E)\}$ for some fixed $E$, and define $T$ and $\ell_E^T$ as in the proof of Proposition~\ref{prop:forwardSSP}. If we define
\[
\hat\theta^{(E,r)} = \argmin_{\theta\in\Theta(E)} -\ell_E^T(\theta; T(Y)) + P_r(\theta),
\]
then 
\[
\hat\theta^{(E,r)} \eqAS \hat\theta^{r} \text{ on } 
A \cap 1_{\{r \leq R_k\}}.
\]
But $\hat\theta^{(E,r)}$ depends only on $T(Y)$. Therefore, on $A$, we can reconstruct the entire path of solutions once we know $T(Y)$.
\end{proof}

@@

\begin{comment}
Sections~\ref{sec:pvalSP}--\ref{sec:modelSSP} discuss sufficient conditions on the $p$-values and the selection algorithm under which single-step $p$-values are automatically independent. Section~\ref{sec:selectionVariables} discusses how we can create independent $p$-values by conditioning on finer selection variables at each step.
